4-3-water-level

### 2022-07-09 1-intro.Rmd
fix abbreviation + more fix on markups

The SpatioTemporal CRAN task view [@ctvspatiotemporal] gathers information about \proglang{R} packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The \pkg{stars} [@stars] package has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with \pkg{sf} [@sf], a package commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualization in \proglang{R}.

@@

The other reason is about tidy data concepts [@tidydata] and how they should be applied to spatio-temporal data. According to the tidy data principles, data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data form given the downstream software such as \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualization. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled, for example in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analyzing spatio-temporal data easier.

@@

The rest of the paper is organized as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualize multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of the \pkg{cubble} package. Section \@ref(others) explains how the \pkg{cubble} package deals with more advanced considerations, including data with hierarchical structure, data matching and how it fits with existing static and interactive visualization tools. Moreover we also illustrate how the \pkg{cubble} package deals with spatio-temporal data transformations. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of the \pkg{cubble} package. An example of how cubble handles NetCDF data is also provided. Section \@ref(conclude) discuss the paper contributions and future directions.

>>>>

The Comprehensive \proglang{R} Archive Network (CRAN) task view SpatioTemporal [@ctvspatiotemporal] gathers information about \proglang{R} packages designed for spatio-temporal data and it has a section on *Representing data* that lists existing spatio-temporal data representations used in \proglang{R}. Among them, @spacetime summarises spatio-temporal data into three forms: time-wide, space-wide, and long formats. The associated package \pkg{spacetime} [@spacetime] implements four spatio-temporal layouts (full grid, sparse grid, irregular, and trajectory) to handle different space and time combinations. The package \pkg{stars} [@stars] has a new implementation to use dense arrays to represent spatio-temporal cubes. It also interfaces with the package \pkg{sf} [@sf], commonly used for wrangling spatial data, and the \pkg{tidyverse} [@tidyverse] suite for general data wrangling and visualization in \proglang{R}.

@@

The other reason is about tidy data concepts [@tidydata] and how they should be applied to spatio-temporal data. According to the tidy data principles, data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data form given the downstream packages such as \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualization. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for hourly, daily or sub-daily periods over years, which are extensively collected and handled, for example in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make data wrangling, visualizing and analyzing spatio-temporal data easier.

@@

The rest of the paper is organized as follows: Section \@ref(cube) introduces the proposed cube structure as a way to conceptualize multivariate spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of the \pkg{cubble} package. Section \@ref(others) explains how the \pkg{cubble} package deals with more advanced considerations, including data with hierarchical structure, data matching and how the package fits with existing static and interactive visualization tools. Moreover we also illustrate how the \pkg{cubble} package deals with spatio-temporal data transformations. Section \@ref(examples) uses Australian weather station data and river level data as examples to demonstrate the use of the package. An example of how the \pkg{cubble} package handles Network Common Data Form (NetCDF) data is also provided. Section \@ref(conclude) discuss the paper contributions and future directions.

### 2022-07-09 3-the-cubble-pkg.Rmd
fix abbreviation + more fix on markups

The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their roles, why the new cubble structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The data \code{climate_flat} is used to illustrate \pkg{cubble} functionality. This is a subset from National Oceanic and Atmospheric Administration (NOAA) Global Historical Climatology Network Daily (GHCND) data. It contains spatial variables, station ID, latitude, longitude, elevation, station name, World Meteorology Organisation ID, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for year 2020. The first five rows of \code{climate_flat} are shown below:

@@

The function \code{as_cubble()} is used to create a cubble with three arguments: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order  (longitude, latitude). The arguments \code{key} and \code{index} follow the wording in the \pkg{tsibble} package to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a cubble out of \code{climate_flat} (a single tibble) with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

Printing a cubble provides some information about the data. Here \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]} and provides information about the coordinates in the data. The third row shows the name and type of all variables nested in the \code{ts} column. In this example, it includes \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

A \code{cubble} object is a subclass of the \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

@@

A cubble can be created from various common spatio-temporal data formats, including basic \proglang{R} objects like \code{tibble}, \code{tsibble}, and \code{sf}. Section \@ref(covid) describes converting multiple tables into a \code{cubble} object and Section \@ref(netcdf) illustrates how to convert a netCDF object.

@@

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and temporarily "hides" the spatial variables. The function \code{face_temporal()} is used to switch from the nested cubble into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the cubble just created into its long form:

@@

The first line in the printed cubble now shows it in the long form and the third line has been changed to display the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built from a \code{grouped_df} class where all the observations from the same site form a group.

```{r face, echo = FALSE, fig.align="center", out.height = "35%", out.width = "100%", fig.cap = "An illustration of function \\code{face\\_temporal} and \\code{face\\_spatial} in cubble. In the first row, \\code{face\\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \\code{face\\_spatial} switches a cubble back to the nested form from the long form and shifts focus back to the spatial aspect."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.002.png"))
```

@@

## Unfold spatial variables into the long cubble {#unfold}

@@

Some readers may question why a new data structure is needed rather than directly creating a list-column  on the combined data using \code{dplyr::nest_by()}. The reason is that cubble is specifically designed to utilize the spatio-temporal structure when arranging observations in a single object. Moreover, it enables easy pivoting between purely spatial, purely temporal, or unfolded into a combined form.

@@

The \code{dplyr} package has many tools for wrangling tidy data,  many of which are useful in the spatio-temporal analysis. The \code{cubble} package provides \code{methods} that support the use of the following \code{dplyr} operations on both the nested and long forms: `mutate`, `filter`, `summarise`, `select`, `arrange`, `rename`, `left_join`, and the slice family (`slice_*`).

@@

A \code{tsibble} object is a tibble where the \code{index} and \code{key} components are used to store temporal and strata information, that makes working with temporal data cognitively efficient. A \code{cubble} object can use a tsibble for storing the temporal information, and effectively utilize the specialist time series operations in the \pkg{tsibble} package. It is easy to cast an existing \code{tsibble} object into a cubble simply by supplying the \code{coords}:

@@

When a nested cubble is created, each element in the list-column \code{ts} has a \code{tsibble} class (labelled \code{tbl_ts})  and operations available to the \code{tsibble} class are still valid on this element. For example, the code below calculates two time series features (mean and variance) of maximum temperature, utilizing the tsibble structure in the cubble:

@@

An \code{sf} object is a spatial data object, containing a feature geometry list-column (\code{sfc}) in the form of various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}). The \pkg{sf} package provides functions that operate efficiently on this spatial information. A cubble can store the spatial information as an \code{sf} object. Methods for the \code{sfc} class can be applied in the nested form of the cubble object.  An illustration is in Section \@ref(covid). The spatial information can also be stored as an \code{s2} in a cubble object.

@@

NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in \proglang{R} exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

Cubble provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble} object. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

@@

NetCDF data can be quite large, and it is sometimes best to subset the data when converting to a cubble. We would recommend reducing to about 300 $\times$ 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution.
The size of spatial grid can be reduced if longer time periods or more variables are needed, through \code{long_range} and \code{lat_range}:

```{r eval = FALSE, echo = TRUE}
dt <- as_cubble(
  my_ncdf, vars = c("q", "z"),
  long_range = seq(-180, 180, 1),lat_range = seq(-90, 90, 1)
  )
```

>>>>

The main functions in the package are \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()}. The following sections explain their roles, why the new \code{cubble} structure is needed and how the package relates to existing packages for spatial and temporal data analysis.

The data \code{climate_flat} is used to illustrate functionality in the \pkg{cubble} package. This is a subset from National Oceanic and Atmospheric Administration (NOAA) [@noaa] Global Historical Climatology Network (GHCN) Daily data. It contains spatial variables, station id, latitude, longitude, elevation, station name, world meteorology organisation id, in additional to daily temporal information,  maximum and minimum temperature values and precipitation records for year 2020. The first five rows of the data are shown below:

@@

The function \code{as_cubble()} is used to create a \code{cubble} object with three arguments: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order (longitude, latitude). The arguments \code{key} and \code{index} follow the wording in the \pkg{tsibble} package to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a \code{cubble} object out of \code{climate_flat} (a single tibble) with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

@@

Printing a \code{cubble} object provides some information about the data. Here \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]} and provides information about the coordinates in the data. The third row shows the name and type of all variables nested in the \code{ts} column. In this example, it includes \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

A \code{cubble} object is a subclass of the \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested \code{cubble}. The rowwise structure makes it simpler to operate on the list using the \code{mutate()} syntax, which is simpler than the \code{purr::map()} when working with a list column. For example, calculating the number of rainy days can be done by:

@@

A \code{cubble} object can be created from various common spatio-temporal data formats, including basic \proglang{R} objects like \code{tibble}, \code{tsibble}, and \code{sf}. Section \@ref(covid) describes converting multiple tables into a \code{cubble} object and Section \@ref(netcdf) illustrates how to convert a NetCDF object.

@@

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). The \code{cubble} class also provides a long form, which expands the \code{ts} column and temporarily "hides" the spatial variables. The function \code{face_temporal()} is used to switch from the nested \code{cubble} into the long one. The first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the \code{cubble} object just created into its long form:

@@

The first line in the header now shows it in the long form and the third line has been changed to display the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long \code{cubble} is built from a \code{grouped_df} class where all the observations from the same site form a group.

```{r face, echo = FALSE, fig.align="center", out.height = "35%", out.width = "100%", fig.cap = "An illustration of function \\code{face\\_temporal} and \\code{face\\_spatial}. In the first row, \\code{face\\_temporal} switches a \\code{cubble} object from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \\code{face\\_spatial} switches a \\code{cubble} object back to the nested form from the long form and shifts focus back to the spatial aspect."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.002.png"))
```

@@

## Unfold spatial variables into the long form {#unfold}

@@

Some readers may question why a new data structure is needed rather than directly creating a list-column  on the combined data using \code{dplyr::nest_by()}. The reason is that the \code{cubble} object is specifically designed to utilize the spatio-temporal structure when arranging observations in a single object. Moreover, it enables easy pivoting between purely spatial, purely temporal, or unfolded into a combined form.

@@

The \pkg{dplyr} package has many tools for wrangling tidy data,  many of which are useful in the spatio-temporal analysis. The \pkg{cubble} package provides methods that support the use of the following operations in the \pkg{dplyr} package on both the nested and long forms: `mutate`, `filter`, `summarise`, `select`, `arrange`, `rename`, `left_join`, and the slice family (`slice_*`).

@@

A \code{tsibble} object is a \code{tibble} where the \code{index} and \code{key} components are used to store temporal and strata information, that makes working with temporal data cognitively efficient. A \code{cubble} object can use a \code{tsibble} to store the temporal information, and effectively utilize the specialist time series operations in the \pkg{tsibble} package. It is easy to cast an existing \code{tsibble} object into a \code{cubble} simply by supplying the argument \code{coords}:

@@

When a nested \code{cubble} is created, each element in the list-column \code{ts} has a \code{tsibble} class (labelled \code{tbl_ts})  and operations available to the \code{tsibble} class are still valid on this element. For example, the code below calculates two time series features (mean and variance) of maximum temperature, utilizing the \code{tsibble} structure in the \code{cubble} object:

@@

An \code{sf} object is a spatial data object, containing a feature geometry list-column (\code{sfc}) in the form of various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}). The package \pkg{sf} provides functions that operate efficiently on this spatial information. A \code{cubble} object can store the spatial information as an \code{sf} object. Methods for the \code{sfc} class can be applied in the nested form of the cubble object.  An illustration is in Section \@ref(covid). The spatial information can also be stored as an \code{s2} vector in a \code{cubble} object.

@@

The NetCDF data is another format commonly used for storing spatio-temporal data. It has two main components: *dimension* for defining the spatio-temporal grid (longitude, latitude, and time) and *variable* that populates the defined grid. Attributes  can be associated with dimensions or variables. Because there can be many different styles of representing this information there is a metadata convention [@hassell_data_2017] to standardize the format of the attributes. A few packages in \proglang{R} exist for manipulating NetCDF data and these include a high-level R interface: \pkg{ncdf4} [@ncdf4], a low-level interface that calls a C-interface: \pkg{RNetCDF} [@rnetcdf], and a tidyverse implementation: \pkg{tidync} [@tidync].

The \pkg{cubble} package provides an \code{as_cubble()} method to coerce the \code{ncdf4} class from the \pkg{ncdf4} package into a \code{cubble} object. It maps each combination of longitude and latitude into an \code{id} as the \code{key}:

@@

Sometimes NetCDF data can be quite large, and it is best to subset the data when converting to a \code{cubble}. We would recommend reducing to about 300 $\times$ 300 grid points for three daily variables in one year. A 300 by 300 spatial grid can be a bounding box of [100, -80, 180, 0] at 0.25 degree resolution or a global bounding box [-180, -90, 180, -90] at 1 degree resolution.
The size of spatial grid can be reduced if longer time periods or more variables are needed, through the arguments \code{long_range} and \code{lat_range}:

```{r eval = FALSE, echo = TRUE}
dt <- as_cubble(
  my_ncdf, vars = c("q", "z"),
  long_range = seq(-180, 180, 1), lat_range = seq(-90, 90, 1)
  )
```

### 2022-07-09 4-other-features.Rmd
fix abbreviation + more fix on markups

Spatial locations can have grouping structures either inherent to the data e.g., state within country or obtained during the analysis e.g., cluster id. In this case, it can be useful to summarize variables at various levels of the hierarchy. The function \code{switch_key()} can be used to change the grouping level of spatial locations. The diagram in Figure \ref{fig:illu-hier} shows how this function can be used to switch the grouping from station ids to cluster ids. The result can also be stretched into long form. By specifying

```{r echo = TRUE, eval = FALSE}
cluster_nested <- station_nested %>% switch_key(key = cluster)
```

the cubble redefines the cubble key from the \code{id} column in \code{station_nested} to the \code{cluster} column in \code{cluster_nested}. All the spatial variables belonging to the \code{cluster} column are now nested into a \code{.val} column which allows for summarizing based on cluster.

@@

One task that may interest spatio-temporal analysts is combining data collected at nearby but not exactly the same sites, for example, weather station measured rainfall and river levels. This can be considered to be a matching problem [@stuart2010matching; @mcintosh2018using] to pair similar time series from nearby locations, or even a data fusion exercise that merges data collected from different sources [@cocchi2019data]. The function \code{match_sites()} in the \pkg{cubble} package provides a simple algorithm for this task. The algorithm first matches spatially by computing the pairwise distance on latitude and longitude. Then it matches temporally by computing the number of matched peaks within a fixed length moving window. Figure \ref{fig:illu-matching} illustrates this temporal matching. In the two series, A and a, three peaks have been identified in each. An interval, of fixed length, is constructed for each peak in series A, while the peaks in series a are tested against whether they fall into any of the intervals. Here two out of three peaks match. Options for \code{match_sites()} are:

@@

The cubble workflow fits works well with an interactive graphics pipeline (e.g., @buja1988elements, @buja1996interactive,  @sutherland2000orca, @xie2014reactive, @cheng2016enabling) that is available in \proglang{R} with the package \pkg{crosstalk} [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series using a cubble. The map (produced from the nested form) and time series (produced from the long form) are both shared \code{crosstalk} objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested \code{cubble} object, which is then communicated to the long cubble -- all the observations with the same id (b) will be selected. The long cubble will then highlight the corresponding series in the time series plot (c).

Linking is also available starting from the time series plot, by selecting points. This will be activate rows having the same id in the long cubble. The corresponding rows in the nested cubble are activated, and highlighted the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what @CS07 would call categorical variable linking, where station id is the categorical variable. <!--\ref{fig:illu-interactive-2} in the Appendix illustrates this process.-->

```{r illu-interactive, echo = FALSE, fig.align="center", out.height = "35%", out.width = "100%", fig.cap = "Linking between multiple plots. The line plots and the map are constructed from shared \\code{crosstalk} objects (long and nested cubbles). When a station is selected on the map (a), the corresponding row in the nested cubble will be activated. This will link to all the rows with the same id in the long cubble (b) and update the line plot (c)."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.004.png"))
```

@@

The package \pkg{GGally} [@ggally] has implemented glyph maps through the \code{glyphs()} function. The function constructs a data frame with calculated position (\code{gx}, \code{gy}, \code{gid}) of each point on the time series using linear algebra (Equations 1 and 2 in @Wickham2012-yr). The data can then be piped into ggplot to create the glyph map as:

@@

Some useful controls over the glyph map are also available in the \code{geom_glyph()} implementation. Polar coordinate glyph maps are specified using \code{polar = TRUE}, and parameters \code{width} and \code{height} can be specified in either absolute or relative value. Global and local scale is specified with \code{global_rescale}, which defaults to \code{TRUE}. Reference boxes and lines can be added with separate \code{geom_glyph_box()} and \code{geom_glyph_line()} lines.

>>>>

Spatial locations can have grouping structures either inherent to the data e.g., state within country or obtained during the analysis e.g., cluster id. In this case, it can be useful to summarize variables at various levels of the hierarchy. The function \code{switch_key()} can be used to change the grouping level of spatial locations. The diagram in Figure \ref{fig:illu-hier} shows how this function can be used to switch the grouping from station ids to cluster ids. The result can also be stretched into long form. By specifying \code{cluster_nested <- station_nested \%>\% switch_key(key = cluster)}, the \code{cubble} objects redefines its \code{key} from the \code{id} column in \code{station_nested} to the \code{cluster} column in \code{cluster_nested}. All the spatial variables belonging to the \code{cluster} column are now nested into a \code{.val} column which allows for summarizing based on cluster.

@@

One task that may interest spatio-temporal analysts is combining data collected at nearby but not exactly the same sites, for example, weather station measured rainfall and river levels. This can be considered to be a matching problem [@stuart2010matching; @mcintosh2018using] to pair similar time series from nearby locations, or even a data fusion exercise that merges data collected from different sources [@cocchi2019data]. The function \code{match_sites()} in the \pkg{cubble} package provides a simple algorithm for this task. The algorithm first matches spatially by computing the pairwise distance on latitude and longitude. Then it matches temporally by computing the number of matched peaks within a fixed length moving window. Figure \ref{fig:illu-matching} illustrates this temporal matching. In the two series, *A* and *a*, three peaks have been identified in each. An interval, of fixed length, is constructed for each peak in series *A*, while the peaks in series *a* are tested against whether they fall into any of the intervals. Here two out of three peaks match. Options for \code{match_sites()} are:

@@

The cubble workflow fits works well with an interactive graphics pipeline (e.g., @buja1988elements, @buja1996interactive,  @sutherland2000orca, @xie2014reactive, @cheng2016enabling) that is available in \proglang{R} with the package \pkg{crosstalk} [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series in a \code{cubble} object. The map (produced from the nested form) and time series (produced from the long form) are both shared \code{crosstalk} objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested \code{cubble} object, which is then communicated to the long \code{cubble} -- all the observations with the same id (b) will be selected. The long \code{cubble} will then highlight the corresponding series in the time series plot (c).

Linking is also available starting from the time series plot, by selecting points. This will be activate rows having the same id in the long \code{cubble} The corresponding rows in the nested \code{cubble} are activated, and highlighted the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what @CS07 would call categorical variable linking, where station id is the categorical variable. <!--\ref{fig:illu-interactive-2} in the Appendix illustrates this process.-->

```{r illu-interactive, echo = FALSE, fig.align="center", out.height = "35%", out.width = "100%", fig.cap = "Linking between multiple plots. The line plots and the map are constructed from shared \\code{crosstalk} objects (long and nested \\code{cubble}). When a station is selected on the map (a), the corresponding row in the nested \\code{cubbles} will be activated. This will link to all the rows with the same id in the long \\code{cubbles} (b) and update the line plot (c)."}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.004.png"))
```

@@

The package \pkg{GGally} [@ggally] has implemented glyph maps through the \code{glyphs()} function. The function constructs a \code{data.frame} with calculated position (\code{gx}, \code{gy}, \code{gid}) of each point on the time series using linear algebra (Equations 1 and 2 in @Wickham2012-yr). The data can then be piped into ggplot to create the glyph map as:

@@

Some useful controls over the glyph map are also available in the \code{geom_glyph()} implementation. Polar coordinate glyph maps are specified using \code{polar = TRUE}, and arguments \code{width} and \code{height} can be specified in either absolute or relative value. Global and local scale is specified with \code{global_rescale}, which defaults to \code{TRUE}. Reference boxes and lines can be added with separate \code{geom_glyph_box()} and \code{geom_glyph_line()} lines.

### 2022-07-09 5-1-covid.Rmd
fix abbreviation + more fix on markups

The five examples here are chosen to illustrate these aspects of using cubble: creating a \code{cubble} object from two COVID data tables with the complication of differing location names, using spatial transformations to make a glyph map of seasonal temperature changes over years, aggregating information spatially to explore precipitation patterns, matching river level data and weather station records for analysis of water supply, reading netCDF format data to reproduce a climate reanalysis plot, and the workflow to create complex interactive linked plots. (There is an additional example and figures in the Appendix, and more examples in the package vignettes.)

>>>>

The five examples here are chosen to illustrate these aspects of the \pkg{cubble} package: creating a \code{cubble} object from two coronavirus (COVID) data tables with the complication of differing location names, using spatial transformations to make a glyph map of seasonal temperature changes over years, aggregating information spatially to explore precipitation patterns, matching river level data and weather station records for analysis of water supply, reading NetCDF format data to reproduce a climate reanalysis plot, and the workflow to create complex interactive linked plots. (There is an additional example and figures in the Appendix, and more examples in the package vignettes.)

### 2022-07-09 5-2-aus-hist-max-temp.Rmd
fix abbreviation + more fix on markups

The Global Historical Climatology Network (GHCN) provides daily climate measures from stations across the world. The data used here (\code{historical_tmax}) is a subset extracted using the package \pkg{rnoaa} [@rnoaa], containing the records of maximum temperature for `r nrow(historical_tmax)` Australian stations from `r range(historical_tmax$first_year)[1]` through `r range(historical_tmax$first_year)[2]` and provides information also on the latitude, longitude and elevation of each of the  stations. <!-- This data  is already cast into a \code{cubble} data object (please see Appendix X for the code to extract  the data from XX and prepare the data into \code{cubble} form), with \code{id} as the key, \code{date} as the index, and \code{c(longitude, latitude)} as the coordinates.--> The goal of this example is to compare the monthly average maximum temperature between two periods, 1971-1975 and 2016-2020, for stations in Victoria and New South Wales, using a glyph map.

First, the stations need to be filtered to those in NSW and Victoria by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by "ASN00") and the next 6 digits encode the station following the Australian Bureau of Meteorology (BOM) [@bom] coding protocols. New South Wales stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a *spatial operation* and hence uses the \code{cubble} object in the nested form:

@@

Next, the monthly maximum average temperature is  calculated for both periods. This is a *temporal operation* requiring a switch into the long cubble form using the \code{face_temporal()} function. In addition, a new indicator for the two time periods of interest is created before the calculation of monthly averages:

@@

```{r basic-manip, out.width="100%", fig.height = 5, fig.width = 10, fig.cap="A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales for the periods (1971-1975, 2016-2020). The corresponding average time series for the Cobar station are display on the top left corner. From the glyph map we can observe that the monthly trend is similar for all locations (low in the winter, high in the summer), and small increased temperatures, particularly in late summer can be seen at most stations in New South Wales."}
tmax <- tmax |> unfold(long, lat)
box_df <- tmax |> face_spatial() |> filter(id == "ASN00048027")
nsw_vic <- ozmaps::abs_ste |> filter(NAME %in% c("Victoria","New South Wales"))

>>>>

The GHCN provides daily climate measures from stations across the world. The data used here (\code{historical_tmax}) is a subset extracted using the package \pkg{rnoaa} [@rnoaa], containing the records of maximum temperature for `r nrow(historical_tmax)` Australian stations from `r range(historical_tmax$first_year)[1]` through `r range(historical_tmax$first_year)[2]` and provides information also on the latitude, longitude and elevation of each of the  stations. <!-- This data  is already cast into a \code{cubble} data object (please see Appendix X for the code to extract  the data from XX and prepare the data into \code{cubble} form), with \code{id} as the key, \code{date} as the index, and \code{c(longitude, latitude)} as the coordinates.--> The goal of this example is to compare the monthly average maximum temperature between two periods, 1971-1975 and 2016-2020, for stations in Victoria and New South Wales (NSW), using a glyph map.

First, the stations need to be filtered to those in Victoria and NSW by using the station identifiers, stored within the 11 digits of the \code{id} variable entries. The country code is in the first 5 digits (Australia is represented by "ASN00") and the next 6 digits encode the station following the Australian Bureau of Meteorology (BOM) [@bom] coding protocols. The NSW stations correspond to entries in the range 46-75 and the Victorian stations to 76-90. Filtering Victoria and New South Wales stations is a *spatial operation* and hence uses the \code{cubble} object in the nested form:

@@

Next, the monthly maximum average temperature is  calculated for both periods. This is a *temporal operation* requiring a switch into the long \code{cubble} form using the \code{face_temporal()} function. In addition, a new indicator for the two time periods of interest is created before the calculation of monthly averages:

@@

```{r basic-manip, out.width="100%", fig.height = 5, fig.width = 10, fig.cap="A glyph map of the monthly maximum average temperature for weather stations in  Victoria and New South Wales (NSW) for the periods (1971-1975, 2016-2020). The corresponding average time series for the cobar station are display on the top left corner. From the glyph map we can observe that the monthly trend is similar for all locations (low in the winter, high in the summer), and small increased temperatures, particularly in late summer can be seen at most stations in NSW."}
tmax <- tmax |> unfold(long, lat)
box_df <- tmax |> face_spatial() |> filter(id == "ASN00048027")
nsw_vic <- ozmaps::abs_ste |> filter(NAME %in% c("Victoria","New South Wales"))

### 2022-07-09 5-3-river-level-rainfall.Rmd
fix abbreviation + more fix on markups

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@bom] in relation to weather station rainfall from NOAA's climate data [@noaa] (\code{climate}). The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using \code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->

>>>>

River level and rainfall data for the same areas should have some similarity. Here we examine the river gauge data (\code{Water_course_level}) from the Bureau of Meteorology [@bom] in relation to weather station rainfall from NOAA's climate data (\code{climate}). The goal is to match water gauges with nearby weather stations, spatially and temporally, using 2020 measurements using \code{match_sites()} function.
<!--The Bureau of Meteorology collects [water data](http://www.bom.gov.au/metadata/catalogue/19115/ANZCW0503900528?template=full) from river gauges. The collected variables include: electrical conductivity, turbidity, watercourse discharge, watercourse level, and water temperature. We expect that the water level should be related to the precipitation available from climate data, since rainfall will raise the water level in the river. Figure \ref{fig:matching-map} shows the location of available weather stations and water gauges in Victoria, and we will try to match water gauges with nearby weather stations, taking into account both spatial and temporal information from 2020.-->
